Variance reduction for MCMC methods via martingale representations

### 2020-04-10 aap/Revision/vr-mcmc-revision-new-new.tex
update

\subsection{Gaussian mixtures}
 We consider  ULA with $\pi$  given by the mixture of two equally-weighted $2$-dimensional Gaussian distributions of the following form
\begin{eqnarray}
\label{eq:gmm}
\pi(x) = \frac{1}{2\sqrt{(2\pi)^{d}|\Sigma|}} \left( \rme^{-\frac{(x-\mu)^T\Sigma^{-1}(x-\mu)}{2}} + \rme^{-\frac{(x+\mu)^T\Sigma^{-1}(x+\mu)}{2}}\right)
\end{eqnarray}
where $\rho \in (0,1), \mu \in \rset^d, \Sigma \in \rset^{2 \times 2}$ and $|\Sigma|$ is its determinant. The function $U(x)$ and its gradient are given by
\[
U(x) = \frac{1}{2}(x-\mu)^T\Sigma^{-1}(x-\mu) - \ln{\left(1 + \rme^{-2\mu^\top \Sigma^{-1}x}\right)}
\]
and
\[
\nabla U(x) = \Sigma^{-1}(x-\mu) +2\mu^\top\Sigma^{-1}\left(1 + \rme^{2 \mu^\top\Sigma^{-1} x}\right)^{-1},
\]
respectively.
In our experiments we take $\mu = \left(0.5,0.5\right)$ and randomly generated positive-definite matrix $\Sigma$ (heterogeneous structure). In order to approximate the expectation \(\pi(f)\) for \(f(x)=\sum_{i=1}^d x_i\) or \(f(x) = \sum_{i=1}^d x^2_i,\) we use ULA with constant step size $\gamma=0.1$ and sampled $\NtrainPath = 10$ independent training trajectories, each one of size $n = 10^4$ to estimate coefficients $a_{p-l,\mathbf{k}}.$ Then we solve the least squares  problems \eqref{eq:Qregr} using the first or second order polynomial basis  depending on $f$. Respectively, we set $K=1$ for linear functionals $f$ and $K=2$ for quadratic functionals. We set the truncation level \(n_{\mathrm{trunc}} = 50\). To test our variance reduction algorithm, we generated 100 independent trajectories of length $n = 10^3$ and depict boxplots of the ergodic averages in Figure ~\ref{fig:GMM}. Our approach (MDCV, Martingale decomposition control variates) is compared to other variance reduction methods of \cite{mira2013zero} and \cite{belomestny2019esvm}. In the baselines we use first order polynomials, as they were used as regressors during computations of $\hat{a}_{p-l,\mathbf{k}}$. Next we compute the cost-variance ratios
\[
\mathcal{R}(K,N,n,n_{\mathrm{trunc}})=\frac{\mathrm{cost}(\pi_{n}^N)\PVar[\pi_{n}^N(f)]}{\mathrm{cost}(\widehat \pi_{K,n}^N) \PVar[\widehat \pi_{K,n,n_{\mathrm{trunc}} }^N(f)|\TrainSet]} 
\]
where variances in the above definition are computed using $24$ independent trajectories of the length $n=2\cdot10^3.$ Note that the cost of $\widehat \pi_{K,n,n_{\mathrm{trunc}}} $ is proportional to the number of nonzero coefficients in \eqref{eq:a-pol} times $n_{\mathrm{trunc}}$. In Figure~\ref{fig:G-costvar} cost-variances ratios are shown for for the case of  Gaussian distribution with parameters. In particular, in Figure~\ref{fig:subfig3}  a function $\mathcal{R}(1,2\cdot 10^3,2\cdot 10^3,n_{\mathrm{trunc}})$ is shown in dependence on $n_{\mathrm{trunc}}$ for the case of a linear functional $f$. In Figure~\ref{fig:subfig4} a similar plot of $\mathcal{R}(2,2\cdot 10^3,2\cdot 10^3,n_{\mathrm{trunc}})$ is shown for the case of a quadratic functional $f.$
Figure~\ref{fig:GMM-costvar} presents the corresponding coast-variance ratios   for a mixture of two two-dimensional normal distributions as in \eqref{eq:gmm} with $\Sigma=I$ and $\mu=(0.5,0.5).$ 
 \begin{figure}[tbh]
\centering
\subfloat[Subfigure 1 list of figures text][$f(x)=\sum_{i=1}^d x_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/GMM_2d_lin_sum.png}
\label{fig:subfig1}}
\qquad
\subfloat[Subfigure 2 list of figures text][$f(x)=\sum_{i=1}^d x^2_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/GMM_2d_lin_sum_squares.png}
\label{fig:subfig2}}
\caption{Boxplots of ergodic averages  from the variance reduced ULA algorithms for the Gaussian mixture model. The compared estimators are the ordinary empirical average  (Vanilla), our estimator of control variates (MDCV), zero variance estimator (ZV) and control variates obtained with empirical spectral variance minimisation (ESVM) 
\label{fig:GMM}}
\end{figure}

\begin{figure}[tbh]
\centering
\subfloat[Subfigure 1 list of figures text][$f(x)=\sum_{i=1}^d x_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_linear_gaussian.png}
\label{fig:subfig3}}
\qquad
\subfloat[Subfigure 2 list of figures text][$f(x)=\sum_{i=1}^d x^2_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_quadratic_gaussian.png}
\label{fig:subfig4}}
\caption{ Cost-variance ratios as a function of the truncation level for  two-dimensional standard Gaussian distribution and different test functions. \label{fig:G-costvar}}
\end{figure}

\begin{figure}[tbh]
\centering
\subfloat[Subfigure 1 list of figures text][$f(x)=\sum_{i=1}^d x_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_linear_gmm.png}
\label{fig:subfig5}}
\qquad
\subfloat[Subfigure 2 list of figures text][$f(x)=\sum_{i=1}^d x^2_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_quadratic_gmm.png}
\label{fig:subfig6}}
\caption{ Cost-variance ratios as a function of the truncation level for  a mixture of two-dimensional  Gaussian distribution and different test functions.
\label{fig:GMM-costvar}}
\end{figure}

>>>>

\subsection{Gaussian mixtures}
 We consider  ULA with $\pi$  given by the mixture of two equally-weighted $2$-dimensional Gaussian distributions of the following form
\label{eq:gmm}
\pi(x) = \frac{1}{2\sqrt{(2\pi)^{d}|\Sigma|}} \left( \rme^{-\frac{(x-\mu)^T\Sigma^{-1}(x-\mu)}{2}} + \rme^{-\frac{(x+\mu)^T\Sigma^{-1}(x+\mu)}{2}}\right)
where $\rho \in (0,1), \mu \in \rset^d, \Sigma \in \rset^{2 \times 2}$ and $|\Sigma|$ is its determinant. The function $U(x)$ and its gradient are given by
U(x) = \frac{1}{2}(x-\mu)^T\Sigma^{-1}(x-\mu) - \ln{\left(1 + \rme^{-2\mu^\top \Sigma^{-1}x}\right)}
\nabla U(x) = \Sigma^{-1}(x-\mu) +2\mu^\top\Sigma^{-1}\left(1 + \rme^{2 \mu^\top\Sigma^{-1} x}\right)^{-1},
respectively.
In our experiments we take $\mu = \left(0.5,0.5\right)$ and randomly generated positive-definite matrix $\Sigma$ (heterogeneous structure). In order to approximate the expectation \(\pi(f)\) for \(f(x)=\sum_{i=1}^d x_i\) or \(f(x) = \sum_{i=1}^d x^2_i,\) we use ULA with constant step size $\gamma=0.1$ and sampled $\NtrainPath = 10^4$ independent training trajectories to estimate coefficients $a_{p-l,\mathbf{k}}.$ Then we solve the least squares  problems \eqref{eq:Qregr} using the first or second order polynomial basis  depending on $f$. Respectively, we set $K=1$ for linear functionals $f$ and $K=2$ for quadratic functionals. We set the truncation level \(n_{\mathrm{trunc}} = 50\). To test our variance reduction algorithm, we generated 100 independent trajectories of length $n = 10^3$ and depict boxplots of the ergodic averages in Figure ~\ref{fig:GMM}. Our approach (MDCV, Martingale decomposition control variates) is compared to other variance reduction methods of \cite{mira2013zero} and \cite{belomestny2019esvm}. In the baselines we use first order polynomials, as they were used as regressors during computations of $\hat{a}_{p-l,\mathbf{k}}$. 
Next we compute the cost-variance ratios
\label{eq:cvR}
\mathcal{R}(K,N,n,n_{\mathrm{trunc}})=\frac{\mathrm{cost}(\pi_{n}^N)\PVar[\pi_{n}^N(f)]}{\mathrm{cost}(\widehat \pi_{K,n}^N) \PVar[\widehat \pi_{K,n,n_{\mathrm{trunc}} }^N(f)|\TrainSet]} 
where variances in the above definition are computed using $24$ independent trajectories each of  length $n=2\cdot10^3.$ Note that the cost of $\widehat \pi_{K,n,n_{\mathrm{trunc}}} $ is proportional to the number of nonzero coefficients in \eqref{eq:a-pol} times $n_{\mathrm{trunc}}$.  In Figure~\ref{fig:G-costvar} cost-variances ratios are shown for for the case of  Gaussian distribution with parameters. In particular, in Figure~\ref{fig:subfig3}  a function $\mathcal{R}(1,2\cdot 10^3,2\cdot 10^3,n_{\mathrm{trunc}})$ is shown in dependence on $n_{\mathrm{trunc}}$ for the case of  linear  $f$. In Figure~\ref{fig:subfig4} a similar plot of $\mathcal{R}(2,2\cdot 10^3,2\cdot 10^3,n_{\mathrm{trunc}})$ is shown for the case of  quadratic $f.$
Figure~\ref{fig:GMM-costvar} presents the corresponding coast-variance ratios   for the mixture of two two-dimensional normal distributions as in \eqref{eq:gmm} with $\Sigma=I$ and $\mu=(0.5,0.5).$  In each case, in order estimate functions $(Q_r),$ we have used $T=10^4$ independent trajectories in the least-squares regression and polynomials of the same order as the order of $f.$
 \begin{figure}[tbh]
\centering
\subfloat[Subfigure 1 list of figures text][$f(x)=\sum_{i=1}^d x_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/GMM_2d_lin_sum.png}
\label{fig:subfig1}}
\qquad
\subfloat[Subfigure 2 list of figures text][$f(x)=\sum_{i=1}^d x^2_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/GMM_2d_lin_sum_squares.png}
\label{fig:subfig2}}
\caption{Boxplots of ergodic averages  from the variance reduced ULA algorithms for the Gaussian mixture model. The compared estimators are the ordinary empirical average  (Vanilla), our estimator of control variates (MDCV), zero variance estimator (ZV) and control variates obtained with empirical spectral variance minimisation (ESVM) 
\label{fig:GMM}}
\end{figure}

\begin{figure}[tbh]
\centering
\subfloat[Subfigure 1 list of figures text][$f(x)=\sum_{i=1}^d x_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_linear_gaussian.png}
\label{fig:subfig3}}
\qquad
\subfloat[Subfigure 2 list of figures text][$f(x)=\sum_{i=1}^d x^2_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_quadratic_gaussian.png}
\label{fig:subfig4}}
\caption{ Cost-variance ratios \eqref{eq:cvR} as  functions of the truncation level $n_{\mathrm{trunc}}$ for  two-dimensional standard Gaussian distribution and different test functions. \label{fig:G-costvar}}
\end{figure}

\begin{figure}[tbh]
\centering
\subfloat[Subfigure 1 list of figures text][$f(x)=\sum_{i=1}^d x_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_linear_gmm.png}
\label{fig:subfig5}}
\qquad
\subfloat[Subfigure 2 list of figures text][$f(x)=\sum_{i=1}^d x^2_i$]{
\includegraphics[width=0.45\textwidth]{../pictures/vr_cost_quadratic_gmm.png}
\label{fig:subfig6}}
\caption{ Cost-variance ratios  \eqref{eq:cvR} as  functions of the truncation level $n_{\mathrm{trunc}}$ for  a mixture of two-dimensional  Gaussian distribution and different test functions.
\label{fig:GMM-costvar}}
\end{figure}
